---
title: "Practical Machine Learning Prediction Project"
author: "Sudev Asok A"
date: "2026-07-14"
output:
  html_document:
    toc: true
    theme: readable
---

# Introduction

This project predicts the manner in which participants performed barbell
lifts using accelerometer measurements collected from the belt, forearm,
arm and dumbbell.

The outcome variable is the `classe` variable.

The data are from the Weight Lifting Exercise Dataset.


# Load Libraries

```{r setup, message=FALSE}

library(caret)
library(randomForest)
library(ggplot2)
library(dplyr)

set.seed(123)

```


# Loading Data


```{r load-data}

training <- read.csv(
  "pml-training.csv",
  na.strings = c("", "NA"),
  stringsAsFactors = FALSE
)


testing <- read.csv(
  "pml-testing.csv",
  na.strings = c("", "NA"),
  stringsAsFactors = FALSE
)


dim(training)

dim(testing)

```


# Data Cleaning


Remove columns containing missing values.


```{r cleaning}


training <- training[
  ,
  colSums(is.na(training)) == 0
]


testing <- testing[
  ,
  colSums(is.na(testing)) == 0
]


```


Remove identification columns.


```{r remove-id}


training <- training[, -c(1:7)]

testing <- testing[, -c(1:7)]


# Convert target variable into factor

training$classe <- as.factor(training$classe)


dim(training)

str(training$classe)


```


# Exploratory Data Analysis


```{r class-distribution}


table(training$classe)


ggplot(
  training,
  aes(x = classe)
) +
geom_bar(
  fill = "steelblue"
) +
theme_minimal()


```


# Training and Validation Split


```{r split-data}


set.seed(123)


inTrain <- createDataPartition(
  training$classe,
  p = 0.75,
  list = FALSE
)


trainData <- training[inTrain, ]

validationData <- training[-inTrain, ]


dim(trainData)

dim(validationData)


```


# Random Forest Model


```{r random-forest}


set.seed(123)


model <- randomForest(
  classe ~ .,
  data = trainData,
  ntree = 100,
  importance = TRUE
)


model


```


# Variable Importance


```{r variable-importance}


varImpPlot(model)


```


# Validation


```{r validation}


validationPrediction <- predict(
  model,
  validationData
)


```


# Confusion Matrix


```{r confusion}


results <- confusionMatrix(
  validationPrediction,
  validationData$classe
)


results


```


# Accuracy


```{r accuracy}


accuracy <- results$overall["Accuracy"]

accuracy


```


# Estimated Out-of-Sample Error


```{r error}


errorRate <- 1 - accuracy

errorRate


```


# Final Prediction


Predict the 20 Coursera test cases.


```{r prediction}


finalPrediction <- predict(
  model,
  testing
)


finalPrediction


```


# Create Submission Files


```{r submission}


dir.create(
  "predictions",
  showWarnings = FALSE
)



for(i in 1:length(finalPrediction))
{
  
  write.table(
    finalPrediction[i],
    file = paste0(
      "predictions/problem_id_",
      i,
      ".txt"
    ),
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE
  )
  
}


```


# Prediction Table


```{r prediction-table}


data.frame(
  Problem = 1:20,
  Prediction = finalPrediction
)


```


# Conclusion


A Random Forest classification model was developed to predict exercise
performance using accelerometer measurements.

The model was evaluated using a validation dataset and achieved high
accuracy. The estimated out-of-sample error was approximately 1%.

The final model was applied to the 20 test cases and prediction files were
created for submission.


# Session Information


```{r session-info}

sessionInfo()

```